 EDA_ Manuscript

### 2025-04-26 code/analysis-code/Statistical_Analysis.qmd
Supp

---
title: "The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women"
author: "Asmith Joseph"
date: "2025-02-23"
output: html_document
---

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine
library(generalhoslem)
library(ResourceSelection)
library(rms)

@@

*Bar Graph of Total Malaria Episodes During Pregnancy by Treatment Arm*

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(here)

# Recode 'total_malaria_episodes_during_pregnancy' into a categorical variable
promo_data_clean <- promo_data_clean %>%
  mutate(
    MalariaEpisodesPreg_cat = case_when(
      total_malaria_episodes_during_pregnancy %in% c(0, 1) ~ "1",
      total_malaria_episodes_during_pregnancy %in% c(2, 3) ~ "2–3",
      total_malaria_episodes_during_pregnancy >= 4 ~ "≥4"
    ),
    MalariaEpisodesPreg_cat = factor(MalariaEpisodesPreg_cat, levels = c("1", "2–3", "≥4"))
  )

# Create the bar graph
ggplot(promo_data_clean, aes(x = MalariaEpisodesPreg_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Total Malaria Episodes During Pregnancy by Treatment Arm",
       x = "Total Malaria Episodes During Pregnancy (Categorical)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

*Bar Graphs for Gravidity and Parity*

```{r}
# Load necessary libraries

# Recode Gravidity into categories: "1", "2–3", "≥4"
promo_data_clean <- promo_data_clean %>%
  mutate(
    Gravidity_cat = case_when(
      gravidity == 1 ~ "1",
      gravidity %in% c(2, 3) ~ "2–3",
      gravidity >= 4 ~ "≥4"
    ),
    Gravidity_cat = factor(Gravidity_cat, levels = c("1", "2–3", "≥4"))
  )

# Create the bar graph for Gravidity by Treatment Arm
ggplot(promo_data_clean, aes(x = Gravidity_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Gravidity Distribution by Treatment Arm",
       x = "Gravidity Category (1, 2–3, ≥4)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

*Parity Distribution by Treatment Arm*

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(here)

# Recode Parity into categories: "0", "1–2", "≥3"
promo_data_clean <- promo_data_clean %>%
  mutate(
    Parity_cat = case_when(
      parity == 0 ~ "0",
      parity %in% c(1, 2) ~ "1–2",
      parity >= 3 ~ "≥3"
    ),
    Parity_cat = factor(Parity_cat, levels = c("0", "1–2", "≥3"))
  )

# Create the bar graph for Parity by Treatment Arm
ggplot(promo_data_clean, aes(x = Parity_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Parity Distribution by Treatment Arm",
       x = "Parity Category (0, 1–2, ≥3)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

@@

*Fit a logistic regression model without the interaction term*

```{r}
# Fit a logistic regression model without the interaction term
model_no_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                              age_at_enrollment_years + gravidity + education_level,
                            family = binomial(link = "logit"),
                            data = promo_data_clean)

# Perform a Likelihood Ratio Test comparing models with and without the interaction
lrt <- anova(model_no_interaction, model_interaction, test = "LRT")
print(lrt)

# Check for multicollinearity using the VIF (Variance Inflation Factor)
library(car)
vif_values <- vif(model_interaction)
print(vif_values)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Generate Analysis of Deviance Table (Likelihood Ratio Test)
anova_out <- anova(model_no_interaction, model_interaction, test = "LRT")

# Convert to a clean data frame
anova_df <- as.data.frame(anova_out)
anova_df$Model <- c("Model 1 (No Interaction)", "Model 2 (With Interaction)")

# Reorder and rename columns for readability
anova_df <- anova_df %>%
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)

colnames(anova_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

# Display table
kable(
  anova_df,
  format = "pandoc",
  caption = "*Table X: Analysis of Deviance Comparing Models With and Without Interaction*"
) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

```{r}
# Fit logistic regression model WITH interaction
model2 <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"),
              data = promo_data_clean)

@@

```{r}
# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

kable(
  vif_df,
  format = "pandoc",
  caption = "*Table Y: Variance Inflation Factors for Model 2 (With Interaction)*"
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

```{r}
# Subset the data to include only women under 25 years of age
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25)

# Create the composite adverse birth outcome variable if not already done
promo_data_young <- promo_data_young %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

# Fit the logistic regression model
model_gravidity <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + study_arm + education_level,
                       family = binomial(link = "logit"),
                       data = promo_data_young)

# Tidy the model output, exponentiating coefficients to yield odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_gravidity)

# Optionally, plot predicted probabilities for gravidity
library(ggeffects)
predicted_probs <- ggeffect(model_gravidity, terms = "gravidity")
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "blue") +
  labs(title = "Predicted Probability of Adverse Outcome by Gravidity (Age < 25)",
       x = "Gravidity",
       y = "Predicted Probability") +
  theme_minimal()

@@

```{r}
# Load required libraries
library(broom)
library(dplyr)
library(knitr)
library(kableExtra)

# Assume model_gravidity is your fitted logistic regression model for women < 25
# For example:
# model_gravidity <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + 
#                          study_arm + education_level,
#                        family = binomial(link = "logit"),
#                        data = promo_data_young)

# Tidy the model output, exponentiating coefficients to get odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

# Create a summary table without the intercept and with neat variable names
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "gravidity" ~ "Gravidity",
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      TRUE ~ term
    ),
    `Odds Ratio (95% CI)` = paste0(round(estimate, 2),
                                   " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  select(term, `Odds Ratio (95% CI)`, p.value) %>%
  rename(Variable = term, `p-value` = p.value)

# Print the table using kable
result_table %>%
  kable(format = "markdown", caption = "")

@@

```{r}
# Tidy the model output and exponentiate coefficients to get odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

# Create a summary table, remove the intercept, and recode variable names
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = dplyr::recode(term,
                         "gravidity" = "Gravidity",
                         "total_malaria_episodes" = "Total Malaria Episodes",
                         "study_armSP" = "Treatment Arm (SP)",
                         "education_levelSecondary" = "Secondary Education",
                         "education_levelTertiary" = "Tertiary Education",
                         "education_levelUniversity" = "University Education"),
    `Odds Ratio (95% CI)` = paste0(round(estimate, 2),
                                   " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  select(term, `Odds Ratio (95% CI)`, p.value) %>%
  rename(Variable = term, `p-value` = p.value)

# Print the table using kable with a descriptive caption
result_table %>%
  kable(format = "markdown", 
        caption = "Adjusted Odds Ratios for Adverse Birth Outcomes Among Young Pregnant Women (<25 Years)")

@@

# Load broom for tidy output
library(broom)

@@

*Train/Test Split and Random Forest Model* which fits a random forest model and plots its ROC curve, is also used for Research Question 1 as an alternative ML approach
```{r}
# 1. Define Random Forest model
rf_model <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# 2. Build the recipe (you already did this correctly)
rf_recipe <- recipe(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

# 3. Combine into a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 4. Tune hyperparameters (optional next step)
set.seed(123)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = vfold_cv(train_data, v = 10, strata = adverse_birth_outcome),
  grid = 20,  # number of hyperparameter combos
  metrics = metric_set(roc_auc)
)

@@

```{r}
# Ensure the outcome is a factor in train_data
train_data <- train_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome))

# Build a recipe for the random forest model (no inline interaction here)
rf_recipe <- recipe(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

# Create a workflow for the random forest model
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# Fit the random forest model on training data
rf_fit <- rf_workflow %>% fit(data = train_data)

# Generate predicted probabilities on the test set
rf_preds <- predict(rf_fit, new_data = test_data, type = "prob")

# Compute ROC for the random forest using the probability for class "1"
library(pROC)
roc_rf <- roc(test_data$adverse_birth_outcome, rf_preds$.pred_1)
plot(roc_rf, col = "red", lwd = 2, main = "ROC Curve: Random Forest")
auc_rf <- auc(roc_rf)
print(paste("Random Forest AUC:", round(auc_rf, 2)))

-   Model Comparison and Interpretation\*

```{r}
# Compare ROC curves
plot(roc_logit, col = "blue", lwd = 2, main = "ROC Comparison: Logistic Regression vs. Random Forest")
lines(roc_rf, col = "red", lwd = 2)
legend("bottomright", legend = c("Logistic Regression", "Random Forest"),
       col = c("blue", "red"), lwd = 2)

@@

```{r}
# Load necessary packages (if not already loaded)
library(tidymodels)
library(xgboost)

# Define a boosting model for classification (using boost_tree)
boost_model <- boost_tree(
  mode = "classification",
  trees = 500,
  learn_rate = tune(),
  tree_depth = tune()
) %>%
  set_engine("xgboost")

>>>>

---
title: "Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting"
author: "Asmith Joseph"
date: "2025-02-23"
output: html_document
---

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine
library(generalhoslem)
library(ResourceSelection)
library(rms)
library(car)

@@

@@

**Model 1: Main‐Effects Logistic Regression (No Interaction)**

```{r compare-models-lrt-vif, message=FALSE, warning=FALSE}
# 1) Subset to the exact variables and drop any rows with missingness
analysis_data <- promo_data_clean %>%
  select(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  ) %>%
  drop_na()

# 2) Fit the model WITHOUT the interaction
model_no_interaction <- glm(
  adverse_birth_outcome ~ 
    total_malaria_episodes +
    study_arm +
    age_at_enrollment_years +
    gravidity +
    education_level,
  family = binomial(link = "logit"),
  data   = analysis_data
)

# 3) Fit the model WITH the regimen × malaria interaction
model_interaction <- glm(
  adverse_birth_outcome ~ 
    total_malaria_episodes * study_arm +
    age_at_enrollment_years +
    gravidity +
    education_level,
  family = binomial(link = "logit"),
  data   = analysis_data
)

# 4) Likelihood–Ratio Test (Analysis of Deviance)
anova_out <- anova(
  model_no_interaction,
  model_interaction,
  test = "LRT"
)

# 5) Tidy LRT into a data frame
anova_df <- as.data.frame(anova_out) %>%
  mutate(
    Model = c("Model 1 (No Interaction)", 
              "Model 2 (With Interaction)")
  ) %>%
  select(
    Model,
    `Resid. Df`,
    `Resid. Dev`,
    Df,
    Deviance,
    `Pr(>Chi)`
  )

# 6) Print the LRT results
print("Analysis of Deviance (LRT) Comparing Models With and Without Interaction:")
print(anova_df)

# 7) Compute VIFs for the interaction model
vif_vals <- car::vif(model_interaction)

if (is.matrix(vif_vals)) {
  # Use the orthonormalized GVIF: GVIF^(1/(2*Df))
  vif_eff <- vif_vals[, "GVIF^(1/(2*Df))"]
  terms   <- rownames(vif_vals)
} else {
  vif_eff <- vif_vals
  terms   <- names(vif_vals)
}

vif_df <- data.frame(
  Term = terms,
  VIF  = vif_eff,
  row.names = NULL
)

# 8) Print the VIF table
print("Variance Inflation Factors for the Interaction Model:")
print(vif_df)

@@

 **Fit Model 2 with the regimen × malaria‐episodes interaction**

```{r vif-model2, message=FALSE, warning=FALSE}
# 1) Fit Model 2 with the regimen × malaria-episodes interaction
model2 <- glm(
  adverse_birth_outcome ~ 
    total_malaria_episodes * study_arm +
    age_at_enrollment_years +
    gravidity +
    education_level,
  family = binomial(link = "logit"),
  data   = promo_data_clean
)

# 2) Compute VIFs treating interaction terms as separate predictors
vif_mat <- vif(model2, type = "predictor")

# 3) Tidy into a data frame
vif_df <- as.data.frame(vif_mat) %>%
  rownames_to_column("Variable") %>%
  rename(
    GVIF           = GVIF,
    `Df (VIF)`     = Df,
    `GVIF^(1/(2*Df))` = `GVIF^(1/(2*Df))`
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes"              ~ "Total malaria episodes",
      Variable == "study_armSP"                         ~ "Treatment arm (SP vs DP)",
      Variable == "age_at_enrollment_years"             ~ "Age at enrollment",
      Variable == "gravidity"                           ~ "Gravidity",
      grepl("^education_level", Variable)               ~ 
        paste("Education:", sub("education_level", "", Variable)),
      Variable == "total_malaria_episodes:study_armSP"  ~ "Interaction: episodes × SP",
      TRUE                                              ~ Variable
    )
  ) %>%
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`)

# 4) Print the VIF table to the console
print(vif_df)

@@

#############################################################################################################################################################################################################

@@

```{r}
# 1) Subset to women <25 and build the composite outcome
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25) %>%
  mutate(
    low_birth_weight      = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 |
                                   stillbirth_bin == 1    |
                                   low_birth_weight   == 1, 1, 0)
  )

# 2) Fit the logistic regression
model_gravidity <- glm(
  adverse_birth_outcome ~ 
    gravidity +
    total_malaria_episodes +
    study_arm +
    education_level,
  family = binomial(link = "logit"),
  data   = promo_data_young
)

# 3) Quick look at the coefficients
summary(model_gravidity)

# 4) Tidy up the output into odds ratios + 95% CI
tidy_model_gravidity <- broom::tidy(
  model_gravidity,
  exponentiate = TRUE,
  conf.int     = TRUE
)
print(tidy_model_gravidity)

# 5) Plot predicted probability vs gravidity
predicted_probs <- ggeffects::ggeffect(
  model_gravidity,
  terms = "gravidity"
)

ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              alpha = 0.2, fill = "steelblue") +
  labs(
    title = "Predicted Probability of Adverse Outcome by Gravidity (Age < 25)",
    x     = "Gravidity (number of prior pregnancies)",
    y     = "Predicted Probability"
  ) +
  theme_minimal()

@@

```{r}
# 2) Subset to women <25 and create the composite outcome
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25) %>%
  mutate(
    low_birth_weight      = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(
      preterm_births_count > 0 |
      stillbirth_bin == 1    |
      low_birth_weight == 1,
      1, 0
    )
  )

# 3) Fit the logistic regression model
model_gravidity <- glm(
  adverse_birth_outcome ~
    gravidity +
    total_malaria_episodes +
    study_arm +
    education_level,
  family = binomial(link = "logit"),
  data   = promo_data_young
)

# 4) Tidy and exponentiate coefficients (to get odds ratios + 95% CIs)
tidy_model_gravidity <- broom::tidy(
  model_gravidity,
  exponentiate = TRUE,
  conf.int     = TRUE
)

# 5) Build a neat summary table (dropping the intercept and renaming terms)
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "gravidity"                 ~ "Gravidity",
      term == "total_malaria_episodes"    ~ "Total Malaria Episodes",
      term == "study_armSP"               ~ "Treatment Arm (SP)",
      term == "education_levelSecondary"  ~ "Secondary Education",
      term == "education_levelTertiary"   ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      TRUE                                ~ term
    ),
    `Odds Ratio (95% CI)` = sprintf(
      "%.2f (%.2f, %.2f)",
      estimate, conf.low, conf.high
    )
  ) %>%
  # first select the raw column names
  dplyr::select(term, `Odds Ratio (95% CI)`, p.value) %>%
  # then rename for presentation
  dplyr::rename(
    Variable = term,
    `p-value` = p.value
  )

# 6) Print the final table
print(result_table)

@@

# Summarize gravidity model: build table of odds ratios (95% CI) and p-values
```{r}
# 1) Subset to <25 and define outcome
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25) %>%
  mutate(
    low_birth_weight      = as.integer(birth_weight < 2.5),
    adverse_birth_outcome = as.integer(
      preterm_births_count > 0 |
      stillbirth_bin == 1    |
      low_birth_weight == 1
    )
  )

# 2) Fit logistic regression
model_gravidity <- glm(
  adverse_birth_outcome ~ 
    gravidity +
    total_malaria_episodes +
    study_arm +
    education_level,
  family = binomial(link = "logit"),
  data   = promo_data_young
)

# 3) Tidy into ORs + 95% CIs
tidy_model_gravidity <- tidy(
  model_gravidity,
  exponentiate = TRUE,
  conf.int     = TRUE
)

# 4) Build a clean summary table
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Variable = case_when(
      term == "gravidity"                ~ "Gravidity",
      term == "total_malaria_episodes"   ~ "Total Malaria Episodes",
      term == "study_armSP"              ~ "Treatment Arm (SP)",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary"  ~ "Tertiary Education",
      term == "education_levelUniversity"~ "University Education",
      TRUE                                ~ term
    ),
    `Odds Ratio (95% CI)` = sprintf(
      "%.2f (%.2f–%.2f)",
      estimate, conf.low, conf.high
    ),
    `p-value` = p.value
  ) %>%
  # explicitly call dplyr::select to avoid masking
  dplyr::select(Variable, `Odds Ratio (95% CI)`, `p-value`)

# 5) Print
print(result_table)

@@

@@

## Random Forest Model Specification and Hyperparameter Tuning
*Train/Test Split and Random Forest Model* which fits a random forest model and plots its ROC curve, is also used for Research Question 1 as an alternative ML approach
```{r}
# 0) Make sure the outcome is a factor
train_data <- train_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0, 1)))

# 1) Define the random‐forest spec
rf_model <- rand_forest(
  trees = 500,
  mtry  = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

library(dials)   # for finalize()

# 2) Build the recipe 
rf_recipe <- recipe(adverse_birth_outcome ~ 
                      total_malaria_episodes + study_arm + 
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

# 3) Assemble workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 4) Extract and finalize parameters
rf_params <- parameters(rf_model) %>%
  finalize(train_data)    # now mtry knows how many predictors

# 5) Create a random tuning grid of size 20
rf_grid <- grid_random(rf_params, size = 20)

# 6) 10-fold stratified CV
rf_folds <- vfold_cv(train_data, v = 10, strata = adverse_birth_outcome)

# 7) Tune
set.seed(123)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = rf_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

# 8) View results
rf_tune %>% 
  collect_metrics() %>% 
  filter(.metric == "roc_auc")

@@

# Random Forest model training and OOB Brier score output
```{r}
# 1)  outcome is a factor
train_data <- train_data %>%
  mutate(
    adverse_birth_outcome = factor(
      adverse_birth_outcome,
      levels = c(0, 1),
      labels = c("no", "yes")
    )
  )

# 2) Define a Random Forest model (with fixed hyperparameters for illustration)
rf_model <- rand_forest(
    trees = 500,
    mtry  = 3,    # you can choose a sensible default or tune later
    min_n = 5     # likewise
  ) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# 3) Build the recipe
rf_recipe <- recipe(
    adverse_birth_outcome ~
      total_malaria_episodes +
      study_arm +
      age_at_enrollment_years +
      gravidity +
      education_level,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors())

# 4) Assemble into a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 5) Fit the random forest
rf_fit <- rf_workflow %>% 
  fit(data = train_data)

# 6) Inspect the fitted workflow
rf_fit

```
- Random forest fit ran without errors, but the OOB Brier score shows it’s essentially no better than random. 
- That echoes what you saw with your logistic and other ML models (AUCs ~0.4–0.5): with these variables alone, none of the algorithms can predict the composite adverse birth outcome.

# Model Comparison and Interpretation
```{r}
# Compare ROC curves
plot(roc_logit, col = "blue", lwd = 2, main = "ROC Comparison: Logistic Regression vs. Random Forest")
lines(roc_rf, col = "red", lwd = 2)
legend("bottomright", legend = c("Logistic Regression", "Random Forest"),
       col = c("blue", "red"), lwd = 2)

@@

# Define a boosting model for classification (using boost_tree)
```{r}
# Define a boosting model for classification (using boost_tree)
boost_model <- boost_tree(
  mode = "classification",
  trees = 500,
  learn_rate = tune(),
  tree_depth = tune()
) %>%
  set_engine("xgboost")

### 2025-04-26 code/eda-code/eda_manuscript.qmd
Supp

```{r}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

library(here)             # Manage file paths
library(dplyr)            # Data manipulation
library(tidyverse)        # Data wrangling and visualization
library(janitor)          # Clean column names
library(skimr)            # Quick data summary
library(lubridate)        # Working with dates
library(readr)            # Reading CSV files
library(ggplot2)          # Visualization
library(ggpubr)           # Publication-ready plots
library(forcats)          # Categorical variables handling

# Table and reporting packages
library(gtsummary)        # Summary tables
library(gt)               # Table formatting
library(knitr)            # Table rendering
library(kableExtra)       # Table customization

# Analysis-specific packages
library(Amelia)           # Missing data visualization
library(pwr)              # Power analysis
library(DiagrammeR)       # Diagrams and flowcharts
library(survival)         # Survival analysis
library(survminer)        # Survival plots
library(ggeffects)        # Marginal effects for regression models
library(broom)            # Tidy model outputs
library(rsample)          # Resampling and splitting datasets

>>>>

```{r}
# Load Required Libraries

# Path management
library(here)               # File path handling

# Data manipulation and wrangling
library(dplyr)              # Data manipulation
library(tidyverse)          # Core tidyverse packages (ggplot2, readr, etc.)
library(janitor)            # Clean column names
library(skimr)              # Quick data summaries
library(lubridate)          # Date/time handling
library(forcats)            # Categorical variable tools

# Visualization
library(ggplot2)            # Data visualization
library(ggpubr)             # Publication-ready plots

# Tables and reporting
library(gtsummary)          # Summary tables
library(gt)                 # Advanced table formatting
library(knitr)              # Report generation
library(kableExtra)         # Enhanced markdown tables

# Data exploration & preparation
library(Amelia)             # Missing-data visualization
library(pwr)                # Power analysis
library(DiagrammeR)         # Diagrams and flowcharts

# Survival analysis
library(survival)           # Survival models
library(survminer)          # Survival plots

# Model effects & outputs
library(ggeffects)          # Marginal effects extraction
library(broom)              # Tidy model outputs

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine
library(generalhoslem)
library(ResourceSelection)
library(rms)
